output:
  github_document:
    math_method: "default"
    md_extensions: +tex_math_dollars
    html_preview: false

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# tvrmst

## Time-Varying Restricted Mean Survival Time from Survival Matrices

`tvrmst` provides **model-agnostic** utilities to compute **restricted mean survival time (RMST)** and **time-varying / conditional RMST** **directly from survival curves on a time grid**.

Unlike model-specific toolkits, `tvrmst` operates on a simple interface:

- a time grid `time` (numeric, strictly increasing)
- a survival object `S`:
  - numeric matrix or data.frame
  - **rows = individuals (units)**, **columns = time points**
  - **require** `ncol(S) == length(time)`

This learner-agnostic input works for:

- Kaplan-Meier curves,
- Cox / parametric model predictions,
- deep survival models (e.g. `survdnn`),
- any estimator that outputs survival probabilities on a grid.


# 1. From RMST to Time-Varying RMST

## 1.1 Classical RMST

For a survival function $S(t)$, the **restricted mean survival time** up to horizon $\tau$ is the area under the survival curve:

$RMST(\tau) = \int_0^{\tau} S(u)\,du$

Interpretation:

> Expected survival time accumulated up to tau (in the same units as time).

This provides an interpretable alternative to hazard ratios, especially under **non-proportional hazards**, delayed effects, or crossing survival curves.


## 1.2 RMST difference curve

For two arms (1 vs 0):

$$\Delta(\tau)=\mathrm{RMST}_1(\tau)-\mathrm{RMST}_0(\tau)$$

Evaluating Delta(tau) over a range of horizons reveals time-dependent patterns such as:

- early harm then late benefit,
- delayed separation,
- sign changes when curves cross.


## 1.3 Window RMST contrast

To localize the effect within a window $[s, s+\tau]$:

$W(s,\tau)=\int_s^{s+\tau} \left(S_1(u)-S_0(u)\right)\,du$


## 1.4 Conditional (landmark) tvRMST

Conditioning on being alive at time $s$:

$\mu_c(s,\tau) = \int_0^{\tau} \frac{S(s+u)}{S(s)}\,du$

Treatment contrast:

$\Delta_c(s,\tau)=\mu_{c,1}(s,\tau)-\mu_{c,0}(s,\tau)$

Interpretation:

> Expected additional survival time in the next tau units given survival up to landmark time s.


# 2. Installation

CRAN version (once available):

```{r, eval=FALSE}
install.packages("tvrmst")
```

Development version:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ielbadisy/tvrmst")
```


# 3. Input Interface

All compute functions use the same strict interface:

* `time`: numeric vector, length `n_time`, strictly increasing
* `S`: numeric matrix or data.frame, `n_units × n_time`

**Required invariant:** `ncol(S) == length(time)` (columns correspond to `time`).

## 3.1 Legacy orientation (optional)

If you have a legacy object where **rows are time** (time × series), use
`as_unit_time()` to convert explicitly:

```{r, eval=FALSE}
S_fixed <- as_unit_time(S_legacy, time, by = "rows")
```

No automatic transposition is performed inside compute functions.

# 4. Minimal API (concise)

| Quantity                               | Function                       | Output         |
| -------------------------------------- | ------------------------------ | -------------- |
| RMST at fixed tau                      | `rmst_tau()`                   | numeric vector |
| RMST curve (tau -> RMST(tau))          | `rmst_curve()`                 | data.frame     |
| Delta RMST curve (Delta(tau))          | `rmst_delta_curve()`           | data.frame     |
| Window contrast W(s, tau)              | `rmst_window()`                | data.frame     |
| Conditional tvRMST mu_c(s, tau)        | `tvrmst_cond()`                | data.frame     |
| Conditional difference Delta_c(s, tau) | `tvrmst_diff()`                | data.frame     |
| Bootstrap CI (KM replicates)           | `bootstrap_tvrmst_diff_reps()` | data.frame     |

Plot helpers (require `ggplot2`):

* `plot_survival_curves()`
* `plot_rmst_curve()`
* `plot_rmst_delta()`
* `plot_tvrmst()`
* `plot_tvrmst_diff()`

# 5. Example 1 : Kaplan-Meier (non-PH illustration)

This example simulates **early harm then late benefit**, illustrating why a single RMST horizon can be misleading.

## Step 1 : Simulate a non-PH trial

```{r}
library(survival)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".")
} else if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".")
} else {
  library(tvrmst)
}

set.seed(20260223)

lambda0 <- 0.085
t_switch <- 6
lambda1_early <- 0.13
lambda1_late  <- 0.02

r_piecewise_exp <- function(n, t_switch, lam_early, lam_late) {
  u <- runif(n)
  H <- -log(u)
  Hs <- lam_early * t_switch
  ifelse(H <= Hs,
         H / lam_early,
         t_switch + (H - Hs) / lam_late)
}

simulate_trial <- function(n_per_arm = 350) {
  T0 <- rexp(n_per_arm, rate = lambda0)
  T1 <- r_piecewise_exp(n_per_arm, t_switch, lambda1_early, lambda1_late)

  C0 <- runif(n_per_arm, 0, 30)
  C1 <- runif(n_per_arm, 0, 30)

  rbind(
    data.frame(arm="Control",
               time=pmin(T0, C0),
               status=as.integer(T0 <= C0)),
    data.frame(arm="Treatment",
               time=pmin(T1, C1),
               status=as.integer(T1 <= C1))
  )
}

df <- simulate_trial()
```

## Step 2 : KM survival on a common time grid

```{r}
t_grid <- seq(0, 24, by = 0.25)

km_on_grid <- function(time, status, t_grid) {
  fit <- survfit(Surv(time, status) ~ 1)
  summary(fit, times = t_grid, extend = TRUE)$surv
}

S0 <- km_on_grid(df$time[df$arm=="Control"],
                 df$status[df$arm=="Control"], t_grid)

S1 <- km_on_grid(df$time[df$arm=="Treatment"],
                 df$status[df$arm=="Treatment"], t_grid)

S0_mat <- rbind(Control   = S0)
S1_mat <- rbind(Treatment = S1)
S_both <- rbind(Control = S0, Treatment = S1)
```

## Step 3 : RMST at two horizons

```{r}
rmst_tau(S_both, t_grid, tau = 8)
rmst_tau(S_both, t_grid, tau = 20)

# two-arm comparison at a fixed horizon
rmst_8 <- rmst_tau(S_both, t_grid, tau = 8)
delta_8 <- unname(rmst_8["Treatment"] - rmst_8["Control"])
delta_8
```

## Step 4 : RMST curve and delta curve (arm comparison)

```{r}
rc <- rmst_curve(S_both, t_grid)
dr <- rmst_delta_curve(rc, arm1 = "Treatment", arm0 = "Control")

# dr$delta is RMST_Treatment(tau) - RMST_Control(tau)
head(dr)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_rmst_delta(
    rmst_delta_df = dr,
    title = "Delta RMST(tau) = Treatment - Control"
  )
}
```

## Step 5 : Conditional tvRMST difference (\Delta_c(s,\tau))

```{r}
s_grid <- seq(0, 18, by = 1)
tau_win <- 6

delta_c <- tvrmst_diff(
  S1 = S1_mat,
  S0 = S0_mat,
  time = t_grid,
  s_grid = s_grid,
  tau = tau_win
)
```

## Step 6 : Plots

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_survival_curves(
    S0 = S0_mat,
    S1 = S1_mat,
    time = t_grid,
    labels = c("Control", "Treatment"),
    title  = "KM survival curves (early harm then late benefit)"
  )

  plot_rmst_delta(
    rmst_delta_df = dr,
    title = "ΔRMST(τ) = RMST_Treatment(τ) - RMST_Control(τ)"
  )

  plot_tvrmst_diff(
    delta_df = delta_c,
    title = paste0("Δc(s, τ) with τ = ", tau_win)
  )
}
```


# 6. Example 2 : Using survdnn predictions (same dataset)

This example shows that `tvrmst` can consume `survdnn` survival predictions **directly**.

## Step 1 : Prepare train/test split

```{r, eval = requireNamespace("survdnn", quietly = TRUE)}
library(survdnn)

df$trt <- as.integer(df$arm == "Treatment")

set.seed(123)
idx_train <- sample(seq_len(nrow(df)), size = floor(0.7 * nrow(df)))

train_data <- df[idx_train, ]
test_data  <- df[-idx_train, ]
```

## Step 2 : Fit a simple SurvDNN

```{r, eval = requireNamespace("survdnn", quietly = TRUE)}
fit_dnn <- survdnn(
  Surv(time, status) ~ trt,
  data = train_data,
  epochs = 50,
  verbose = FALSE
)
```

## Step 3 : Predict survival curves on the same grid

`survdnn` requires explicit `times` for `type="survival"`.

```{r, eval = requireNamespace("survdnn", quietly = TRUE)}
pred_df <- predict(
  fit_dnn,
  newdata = test_data,
  type = "survival",
  times = t_grid
)
```

`pred_df` is a data.frame with rows = individuals and columns = time points.
`tvrmst` accepts this directly when you pass `time = t_grid`.

## Step 3.5 : Individualized predictions (CoxPH, lightweight)

Use a simple Cox model with a heterogeneous covariate to ensure individual-level
variation in predicted curves.

```{r, eval = requireNamespace("survival", quietly = TRUE)}
df$trt <- as.integer(df$arm == "Treatment")
set.seed(1)
df$x <- rnorm(nrow(df))   # ensures individual heterogeneity

cox_fit <- survival::coxph(survival::Surv(time, status) ~ trt + x, data = df)

time_grid <- t_grid
sf <- survival::survfit(cox_fit, newdata = df)

# survfit summary returns time × n, transpose to n × time (unit-first)
S_ind <- t(summary(sf, times = time_grid, extend = TRUE)$surv)

# split by arm (unit-first)
S0_units <- S_ind[df$arm == "Control", , drop = FALSE]
S1_units <- S_ind[df$arm == "Treatment", , drop = FALSE]

stopifnot(ncol(S_ind) == length(time_grid))
stopifnot(nrow(unique(round(S_ind[1:50, 1:50], 6))) > 1)
```

## Step 4 : Conditional tvRMST per individual

```{r, eval = requireNamespace("survdnn", quietly = TRUE)}
mu_indiv <- tvrmst_cond(
  S = pred_df,                 # n_units × n_time data.frame
  time = t_grid,
  s_grid = seq(0, 12, by = 2),
  tau = 6
)
```

## Step 5 : Aggregate across individuals

```{r, eval = requireNamespace("survdnn", quietly = TRUE)}
mu_mean <- rowMeans(mu_indiv[, -1])
out <- data.frame(s = mu_indiv$s, mean_tvRMST = mu_mean)
out
```

## Step 6 : Individualized RMST curves

```{r, eval = requireNamespace("survival", quietly = TRUE)}
RMST_mat <- rmst_dynamic(S_ind, time_grid)

unit_n <- ncol(RMST_mat) - 1
rmst_long <- data.frame(
  id  = rep(seq_len(unit_n), each = length(time_grid)),
  tau = rep(time_grid, times = unit_n),
  rmst = as.vector(as.matrix(RMST_mat[, -1, drop = FALSE])),
  arm  = rep(df$arm, each = length(time_grid))
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(rmst_long, ggplot2::aes(x = tau, y = rmst, group = id)) +
    ggplot2::geom_line(alpha = 0.1) +
    ggplot2::facet_wrap(~ arm)
}
```


# 7. Bootstrap confidence intervals

## 7.1 Bootstrap from KM replicates (recommended)

Resample individuals within each arm, recompute KM, and propagate uncertainty into $\Delta_c(s,\tau)$.

```{r}
B <- 200

ctrl_idx <- which(df$arm == "Control")
trt_idx  <- which(df$arm == "Treatment")

make_rep <- function() {
  d0 <- df[sample(ctrl_idx, replace = TRUE), ]
  d1 <- df[sample(trt_idx,  replace = TRUE), ]

  S0b <- km_on_grid(d0$time, d0$status, t_grid)
  S1b <- km_on_grid(d1$time, d1$status, t_grid)

  list(
    S0 = rbind(Control   = S0b),
    S1 = rbind(Treatment = S1b)
  )
}

reps <- lapply(seq_len(B), function(i) make_rep())

boot_reps <- bootstrap_tvrmst_diff_reps(
  reps = reps,
  time = t_grid,
  s_grid = s_grid,
  tau = tau_win,
  conf = 0.95
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_tvrmst_diff(
    delta_df = boot_reps,
    title = paste0("Bootstrap CI for Δc(s, τ), τ = ", tau_win)
  )
}
```

Row bootstrap has been removed from the package to avoid degenerate usage. For trial-style
inference, use `bootstrap_tvrmst_diff_reps()` (Section 7.1).


# 8. Identity check (internal consistency)

`tvrmst` provides an algebraic identity check:

$\mathrm{RMST}(s+\tau)-\mathrm{RMST}(s) = S(s)\,\mu_c(s,\tau)$

```{r}
# check_identities(S, time, s_grid, tau, eps = ...)
```


# Citation

If you use `tvrmst`, please cite the package paper (R Journal submission):

> EL BADISY, I. (2026). *tvrmst: Time-Varying RMST from Survival Matrices*. The R Journal (submitted).


# License

MIT.


## References (conceptual background)

* Royston, P. & Parmar, M. K. B. (2011). RMST as an alternative to the hazard ratio under non-proportional hazards.
* Huang, B. et al. (2020). RMST-based estimation as interpretable effect size under non-PH.
* Syriopoulou, E. et al. (2022). Absolute survival contrasts for interpretability.
* Yang, S. et al. (2021). Dynamic/conditional RMST formulations in landmarking/regression frameworks.
